Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking

### 2023-07-13 manuscript/manuscript.qmd
fix minor errors in manus text.

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    pdf-engine: lualatex
    keep-tex: false
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  simple-article-pdf:
    indent: false
    classoption: 
     - 9pt
     - lucida
     - twocolumn
     - twoside
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

Body-worn accelerometers have emerged as an effective and affordable
alternative for objectively assessing sleep patterns at home over
extended periods. These devices gather continuous, high-resolution data
for several weeks without the need for recharging, thus reducing
participant burden. Initial applications of accelerometry for sleep and
wake stage classification were based on wrist movements, starting with
an algorithm developed in 1982 and validated with PSG [@webster1982].
This model was later refined in 1992 [@cole1992], giving rise to the
widely used Cole-Kripke model. As the field advanced, an array of
techniques, including heuristic algorithms, machine learning models,
regression, and deep learning, were employed to analyze wrist-worn
accelerometer data[@palotti2019; @cole1992; @sazonov2004; @sadeh1994;
@hees2015; @sundararajan2021].

While wrist and hip-worn devices have benefited from extensive
methodological development, thigh-worn accelerometers have not seen the
same level of advancement. Existing studies mainly focus on
distinguishing sleep from wakefulness, with emphasis on defining 'waking
time' and 'bedtime' [@carlson2021; @inan-eroglu2021; @vanderberg2016;
@winkler2016]. Recent strides in estimating sleep duration using these
devices have been made, including the introduction of a promising
algorithm and its comparison against PSG[@johansson_development_2023].
Despite these advancements, the application of machine learning
techniques in this area is still relatively unexplored. Considering the
potential of thigh-worn accelerometers for accurate physical behavior
assessment[@skotte_detection_2014; @arvidsson2019], there is a
significant research gap. Therefore, future studies need to develop
techniques similar to those used for wrist and hip-worn accelerometers,
with the ultimate goal of establishing a more holistic, accurate, and
user-friendly method of sleep and physical activity tracking.

@@

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we
considered the gold standard for measuring sleep. Furthermore, our
secondary goal was to assess the developed models' performance in
evaluating important sleep quality parameters, including sleep period
time (SPT), total sleep time (TST), sleep efficiency (SE), latency until
persistent sleep (LPS), and wake after sleep onset (WASO).

@@

The current study leverages data from the SCREENS
project[@rasmussen2020], a study conducted from October 2018 to March
2019 in Middelfart, Southern Denmark, that evaluated the impact of
screen media usage on Danish families. For our analysis, we focused on
data from child participants aged between 6 and 10 years within the
SCREENS cohort. Our primary sources of data were accelerometer readings
from Axivity AX3 devices attached to the children's thighs, and
electroencephalography (EEG) data derived from the ZM device. The
Axivity AX3, an unobtrusive 3-axis accelerometer, was positioned midway
between the hip and knee on the right anterior thigh, recording
participant movement data.

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, which utilizes advanced EEG hardware and
signal processing algorithms, employs three self-adhesive, disposable
sensors placed outside the hairline for reliable EEG signal acquisition.
The ZM uses two proprietary algorithms: Z-ALG and Z-PLUS. The Z-ALG is
utilized for accurate sleep detection, showcasing its suitability for
in-home monitoring[@kaplan2014], while the Z-PLUS effectively
differentiates sleep stages, as evidenced by its alignment with expert
evaluations using PSG data[@wang2015]. In the current study, we treated
all sleep stages as a single category effectively deducing the output of
the ZM to "awake" and "asleep" as the ability to distinguish sleep
stages are not a necessity to derive sleep quality parameters of
interest and to simplify the learning process of the models.

@fig-flow illustrates the selection criteria applied to the children's
recordings from the SCREENS study. We included only ZM recordings that
were accompanied by complete accelerometer data and lasted between 7 and
14 hours. Any night when the ZM reported sensor issues was excluded. The
children whose recordings were considered had an average age of 9.4
years, with a standard deviation of 2.1. In their raw form, the ZM
predictions encompassed 696,779 epochs, each 30 seconds long. Notably,
approximately 84% of the total ZM recording duration was classified as
sleep, resulting in an imbalance of the dataset.

@@

In this study, data processing of the raw accelerometer data began with
a low-pass filtration step using a 4th order Butterworth filter with a 5
Hz cut-off frequency to eliminate high-frequency noise. Following
filtration, data were partitioned into overlapping 2-second intervals,
each successive interval sharing a 50% overlap with the previous one
similar to methods described by Skotte et al.[@skotte_detection_2014].
Any non-wear data was remove using previously described
methods[@skovgaard2023] and data was resampled to 30-second epochs so
every sample classified by the algorithms corresponds to a 30-second
epoch scored during the ZM recordings. Subsequently, we performed a
feature extraction process that yielded a set of 88 features, providing
a robust characterization of the data. Extracted from accelerometer and
temperature signals, these features include temporal elements that use
both lag and lead values, capturing dynamic data trends by incorporating
measurements from preceding and upcoming intervals. Furthermore,
inspired by Walch et al.[@walch2019], we incorporated sensor-independent
features to encapsulate circadian rhythms. These features offer unique
insights not directly discernible from sensor outputs and are meant to
approximate the changing drive of the circadian clock to sleep over the
course of the night (see @fig-sensor-independent). Furthermore, the
feature set was enriched by including signal characteristics, which
encompass vector magnitude, mean crossing rate, skewness, and kurtosis
for each of the x, y, and z dimensions. All features are summarized in
table ??? **in the supplementary matrials**. Subsequently, we merged the
ZM and corresponding accelerometer recordings. Any overlapping time
between the ZM and accelerometer data was treated as 'in-bed' time, with
the remaining time considered 'out-of-bed'. This process yielded a
dataset providing a around the clock temporal view of each participant's
activity and sleep patterns.

@@

In addition to the engineered features, we chose to incorporate the
median-filtered raw predictions from the ZM device into our modeling
process. This decision stemmed from the understanding that children
typically undergo around five to eight sleep cycles per night, with
awakenings most likely occurring at the end of each
cycle[@galland_normal_2012]. Upon examining the raw ZM predictions, we
noted a significant overestimation in the number of awakenings per night
for the children in our study, exceeding what would be expected based on
typical sleep cycle patterns. In particular, many of these brief
awakenings could be considered as noise, which when present in the data,
can potentially hinder the learning process of machine learning models
by obscuring the underlying patterns that the models are trying to
learn, leading to less accurate predictions. Consequently, we elected to
train and evaluate our models using not only the raw ZM output, but also
versions that were subjected to 5-minute and 10-minute median filters.
This approach, by mitigating this noise, resulted in an anticipated,
more age-appropriate count of awakenings per night, providing a more
accurate depiction of children's sleep patterns (see @fig-zm-median).

@@

In this study, we employed two distinct modeling strategies to analyze
sleep patterns from thigh-mounted accelerometer data. We used a
sequential strategy, comprising an ensemble of four pairs of models,
each pair featuring the same algorithm. This strategy aimed to make the
prediction task more manageable for the algorithms by breaking it down
into a sequence of two binary classifications: first predicting 'in-bed'
time, then 'sleep' time. Simultaneously, we also used a multiclass
approach utilizing a bidirectional Long Short-Term Memory
(biLSTM)[@hochreiter1997] neural network.

@@

For the models in sequence, we trained four pairs of classification
models. Each pair was designed to distinguish between in-bed/out-of-bed
and asleep/awake states, respectively. The dataset was randomly split
into a training set and a testing set, each containing approximately 50%
of the subjects. This division ensured that samples from the same night
were never simultaneously present in both sets. To optimize
hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
a regular grid, comprising 20 different combinations of hyperparameters.
The F1 score served as the optimization metric. The best-performing set
of hyperparameters was then used to fit the models to the full training
dataset. This approach allowed us to maximize performance by leveraging
all available data. Moreover, after extracting the in-bed time from the
initial sequential models, the imbalance on the resulting dataset could
cause biases during model training, as models may favor predicting the
majority class. To rectify this, we employed the Synthetic Minority
Over-sampling Technique (SMOTE)[@chawla2002]. SMOTE generates new
samples by interpolating random samples with their nearest neighbors. We
utilized the themis R package[@themis] to implement SMOTE, resulting in
a balanced distribution of training samples across both classes.

@@

Our study's XGBoost model demonstrated relatively narrower limits of
agreements (LOAs) for TST, SE, and WASO, with ranges of -95.5 to 81.4
min, -15.6 to 13.3%, and -83.4 to 81.7 min, respectively when compared
with other models such as the Van Hees algorithm[@hees2015], Oakley rsc
(rescored)[@palotti2019], and LSTM-50[@palotti2019] evaluated in the
Patterson et al. study[@patterson_40_2023]. Furthermore, comparing the
LOAs between our XGBoost model and the algorithm developed for
thigh-worn devices by Johansson et al. study[@johansson2022], our
XGBoost model showed narrower LOAs for TST , SE, LPS , and WASO, but not
SPT. Generally, all methods, both from this study and from the reviewed
literature, exibit wide LOAs sugesting that there is high variability in
the derevide sleep quality parameters. These findings imply that the
current methods, are only reasonbly reliable for assessing sleep quality
parameters at a group level. However, caution should be exercised when
applying the models and methods to individual-level sleep assessments.
Therefore, further improvements and refinements are needed to enhance
the precision and reliability of these models for individual sleep
assessments.

>>>>

---
title: "Improving Sleep Quality Estimation: A Comparative Study of Machine Learning and Deep Learning Techniques Utilizing Free-Living Accelerometer Data from Thigh-Worn Devices and EEG-Based Sleep Tracking"
author:
  - name: Esben Høegholm Lykke
    email: eskovgaard@health.sdu.dk
    affiliations: 
        - id: some-tech
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
    attributes:
        corresponding: true
  - name: Jan Christian Brønd
    email: jbrond@health.sdu.dk
    affiliations:
        - id: another-u
          name: University of Southern Denmark
          department: Department of Sports Science and Clinical Biomechanics
          address: Campusvej 55
          city: Odense
          state: Denmark
          postal-code: 5230
abstract: "LÆS IKKE! Det er volapyk på latin. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse vitae dictum eros, ullamcorper elementum orci. Sed laoreet nulla neque, pulvinar fermentum mi iaculis at. Nulla ultricies nibh sit amet vestibulum rutrum. Nam pharetra nisl sed ipsum maximus suscipit. Duis metus nunc, ullamcorper eu mi rutrum, tempus ultricies ante. Nunc vitae lectus nisi. Aliquam efficitur ut eros ut pellentesque. Aenean blandit, nisl nec efficitur interdum, nisi ipsum fermentum dolor, at tempus sem turpis in lacus. Curabitur sollicitudin lectus sit amet velit pellentesque laoreet. Ut posuere diam lobortis nisi eleifend tincidunt. Ut at euismod sem, sed dignissim ligula. Aliquam lacinia massa libero, id eleifend velit pulvinar ac. Fusce volutpat elit eu nulla viverra, nec tempus orci pellentesque."
keywords: 
  - Sleep
  - Accelerometry
  - EEG
  - Machine learning
  - Sleep quality
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: false
    journal:
      name: npj Digital Medicine
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header: 
      text: '\usepackage{lineno}\linenumbers'
    fig-pos: 'b'
  docx:
    number-sections: true
editor: 
  markdown: 
    wrap: 72
editor_options: 
  chunk_output_type: inline
---

@@

Body-worn accelerometers have emerged as an effective and affordable
alternative for objectively assessing sleep patterns at home over
extended periods. These devices gather continuous, high-resolution data
for several weeks without the need for recharging, thus reducing
participant burden. Initial applications of accelerometry for sleep and
wake stage classification were based on wrist movements, starting with
an algorithm developed in 1982 and validated with PSG [@webster1982].
This model was later refined in 1992 [@cole1992], giving rise to the
widely used Cole-Kripke model. As the field advanced, an array of
techniques, including heuristic algorithms, machine learning models,
regression, and deep learning, were employed to analyze hip
andwrist-worn accelerometer data[@palotti2019; @cole1992; @sazonov2004;
@sadeh1994; @hees2015; @sundararajan2021].

While wrist and hip-worn devices have benefited from extensive
methodological development, thigh-worn accelerometers have not seen the
same level of advancement. Existing studies mainly focus on
distinguishing sleep from wakefulness, with emphasis on defining 'waking
time' and 'bedtime' [@carlson2021; @inan-eroglu2021; @vanderberg2016;
@winkler2016]. Recent strides in estimating sleep duration using these
devices have been made, including the introduction of a promising
algorithm and its comparison against PSG[@johansson_development_2023].
Despite these advancements, the application of machine learning
techniques in this area is still unexplored. Considering the potential
of thigh-worn accelerometers for accurate physical behavior
assessment[@skotte_detection_2014; @arvidsson2019], there is a
significant research gap. Therefore, future studies need to develop
techniques similar to those used for wrist and hip-worn accelerometers,
with the ultimate goal of establishing a more holistic, accurate, and
user-friendly method of sleep and physical activity tracking.

@@

Our primary objective in this study was to evaluate a range of machine
learning and deep learning models, utilizing the raw data collected from
a tri-axial thigh-worn accelerometer to estimate in-bed and sleep time.
To ensure the reliability and effectiveness of our models, we compared
their outputs with an EEG-based sleep tracking device, which we, in this
current study, considered as the gold standard for measuring sleep.
Furthermore, our secondary goal was to assess the developed models'
performance in evaluating important sleep quality parameters, including
sleep period time (SPT), total sleep time (TST), sleep efficiency (SE),
latency until persistent sleep (LPS), and wake after sleep onset (WASO).

@@

The current study leverages data from the SCREENS
project[@rasmussen2020], a study conducted from October 2018 to March
2019 in Middelfart, Southern Denmark, that evaluated the impact of
screen media usage on Danish families. For our analysis, we focused on
data from child participants aged between 6 and 10 years within the
SCREENS cohort. Our primary sources of data were accelerometer readings
from Axivity AX3 devices attached to the children's thighs, and
electroencephalography data derived from the ZM device. The Axivity AX3,
an unobtrusive 3-axis accelerometer, was positioned midway between the
hip and knee on the right anterior thigh, recording participant movement
data.

Sleep state information was extracted using the ZM, a product of General
Sleep Corporation. The ZM, which utilizes advanced EEG hardware and
signal processing algorithms, employs three self-adhesive, disposable
sensors placed outside the hairline for reliable EEG signal acquisition.
The participants of the SCREENS study were instructed to attach the
device when they went to bed. The ZM uses two proprietary algorithms:
Z-ALG and Z-PLUS. The Z-ALG is utilized for accurate sleep detection,
showcasing its suitability for in-home monitoring[@kaplan2014], while
the Z-PLUS effectively differentiates sleep stages, as evidenced by its
alignment with expert evaluations using PSG data[@wang2015]. In the
current study, we treated all sleep stages as a single category
effectively deducing the output of the ZM to "awake" and "asleep" as the
ability to distinguish sleep stages are not a necessity to derive the
sleep quality parameters of interest and to simplify the learning
process of the models.

@fig-flow illustrates the selection criteria applied to the children's
recordings from the SCREENS study. We included only ZM recordings that
were accompanied by complete accelerometer data and lasted between 7 and
14 hours. Any night when the ZM reported sensor issues was excluded
yielding 585 nights included in the study. The children whose recordings
were considered had an average age of 9.4 years, with a standard
deviation of 2.1. In their raw form, the ZM predictions encompassed
696,779 epochs, each 30 seconds long. Notably, approximately 84% of the
total ZM recording duration was classified as sleep, resulting in an
imbalance of the ZM dataset.

@@

In this study, data processing of the raw accelerometer data began with
a low-pass filtration step using a 4th order Butterworth filter with a 5
Hz cut-off frequency to eliminate high-frequency noise. Following
filtration, data were partitioned into overlapping 2-second intervals,
each successive interval sharing a 50% overlap with the previous one
similar to methods described by Skotte et al.[@skotte_detection_2014].
Any non-wear data was removed using previously described
methods[@skovgaard2023] and data was resampled to 30-second epochs so
every sample classified by the algorithms corresponds to a 30-second
epoch scored during the ZM recordings. Subsequently, we performed a
feature extraction process that yielded a set of 88 features, providing
a robust characterization of the data. Extracted from accelerometer and
temperature signals, these features include temporal elements that use
both lag and lead values, capturing dynamic data trends by incorporating
measurements from preceding and upcoming epochs. Furthermore, inspired
by Walch et al.[@walch2019], we incorporated sensor-independent features
to encapsulate circadian rhythms. These features offer unique insights
not directly discernible from sensor outputs and are meant to
approximate the changing drive of the circadian clock to sleep over the
course of the night (see @fig-sensor-independent). Furthermore, the
feature set was enriched by including signal characteristics, which
encompass vector magnitude, mean crossing rate, skewness, and kurtosis
for each of the x, y, and z dimensions. All features are summarized in
table ??? **in the supplementary matrials**. Subsequently, we merged the
ZM and corresponding accelerometer recordings. Any overlapping time
between the ZM and accelerometer data was treated as 'in-bed' time, with
the remaining time considered 'out-of-bed'. This process yielded a
dataset providing a around the clock temporal view of each participant's
activity and sleep patterns.

@@

In addition to the engineered features, we chose to incorporate the
median-filtered raw predictions from the ZM device into our modeling
process. This decision stemmed from the understanding that children
typically undergo around five to eight sleep cycles per night, with
awakenings most likely occurring at the end of each
cycle[@galland_normal_2012]. Upon examining the raw ZM predictions, we
noted a significant overestimation in the number of awakenings per night
for the children in our study, exceeding what would be expected based on
typical sleep cycle patterns (see @fig-zm-median). In particular, many
of these brief awakenings could be considered as noise, which when
present in the data, can potentially hinder the learning process of
machine learning models by obscuring the underlying patterns that the
models are trying to learn, leading to less accurate predictions.
Consequently, we elected to train and evaluate our models using not only
the raw ZM output, but also versions that were subjected to 5-minute and
10-minute median filters. This approach, by mitigating this noise,
resulted in an anticipated, more age-appropriate count of awakenings per
night, providing a more accurate depiction of children's sleep patterns
(see @tbl-zm_overview).

@@

In this study, we employed two distinct learning strategies to analyze
sleep patterns from thigh-mounted accelerometer data. We used a
sequential strategy, comprising an ensemble of four pairs of models,
each pair featuring the same algorithm. This strategy aimed to make the
prediction task more manageable for the algorithms by breaking it down
into a sequence of two binary classifications: first predicting 'in-bed'
time, then 'sleep' time. Simultaneously, we also used a multiclass
approach utilizing a bidirectional Long Short-Term Memory
(biLSTM)[@hochreiter1997] neural network.

@@

For the models in sequence, we trained four pairs of classification
models. Each pair was designed to distinguish between in-bed/out-of-bed
and asleep/awake states, respectively. The dataset was randomly split
into a training set and a testing set, each containing approximately 50%
of the subjects. This division was ensured to not have samples from the
same night simultaneously present in both sets. To optimize
hyperparameters, we performed a 10-fold Monte Carlo cross-validation on
a regular grid, comprising 20 different combinations of hyperparameters.
The F1 score served as the optimization metric. The best-performing set
of hyperparameters was then used to fit the models to the full training
dataset. This approach allowed us to maximize performance by leveraging
all available data. Moreover, after extracting the in-bed time from the
initial sequential models, the imbalance on the resulting dataset could
cause biases during model training, as models may favor predicting the
majority class. To rectify this, we employed the Synthetic Minority
Over-sampling Technique (SMOTE)[@chawla2002]. SMOTE generates new
samples by interpolating random samples with their nearest neighbors. We
utilized the themis R package[@themis] to implement SMOTE, resulting in
a balanced distribution of training samples across both classes.

@@

Our study's XGBoost model demonstrated relatively narrower limits of
agreements (LOAs) for TST, SE, and WASO, with ranges of -95.5 to 81.4
min, -15.6 to 13.3%, and -83.4 to 81.7 min, respectively when compared
with other models such as the Van Hees algorithm[@hees2015], Oakley rsc
(rescored)[@palotti2019], and LSTM-50[@palotti2019] evaluated in the
Patterson et al. study[@patterson_40_2023]. Furthermore, comparing the
LOAs between our XGBoost model and the algorithm developed for
thigh-worn devices by Johansson et al.
study[@johansson_development_2023], our XGBoost model showed narrower
LOAs for TST , SE, LPS , and WASO, but not SPT. Generally, all methods,
both from this study and from the reviewed literature, exibit wide LOAs
sugesting that there is high variability in the derevide sleep quality
parameters. These findings imply that the current methods, are only
reasonbly reliable for assessing sleep quality parameters at a group
level. However, caution should be exercised when applying the models and
methods to individual-level sleep assessments. Therefore, further
improvements and refinements are needed to enhance the precision and
reliability of these models for individual sleep assessments.
